Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-06-01 submission/Sze_Obesity_PeerJ_2016.Rmd
Work through Results section

# Lack of power and small effect size confounds the ability to differentiate obese and non-obese individuals using gut microbiome data

@@

```{r datasets}
alpha <- read.table(file="data/process/alpha_composite.summary", header=T)
shannon_effect_size <- 100*abs(1-alpha[alpha$metric=="shannon","obese"]/alpha[alpha$metric=="shannon","non_obese"])
beta <- read.table(file="data/process/beta_tests.summary", header=T)

genus <- read.table(file="data/process/random_forest.genus.summary", header=T)
otu <- read.table(file="data/process/random_forest.otu.summary", header=T)

@@

@@

## Abstract
Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity. This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets. The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis. Alpha-diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances. Although these tests yielded small P-values, the difference between the Shannon diversity index of obese and non-obese individuals was `r shannon_effect_size`%. A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity. When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`). Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.

@@

There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans. The first such study used genetically obese mice and observed the ratio of the relative abundances of Bacteroidetes to Firmicutes (B:F) was lower in obese mice than lean mice [@ley_obesity_2005]. Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009]. They also showed that the relative abundance of Bacteroidetes and Firmicutes increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006]. Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since different communities yield variation in animal weight [REFERENCES]. The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.

@@

***Separate and Pooled Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B:F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity [Figure 2A], three for observed richness [Figure S1], one for evenness [Figure S2], and one for Bacteroidetes [Figure S3]. There were no significant results for Firmicutes [Figure S4] or the B:F ratio [Figure 2B]. Using the linear mixed-effect model to analyze the data pooled together while also controling study effects we found that for Shannon diversity  there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 2A]. Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`) [Figure S1].  However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B:F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`)  [Figure 2B & S2-S4] using this approach, between obese and non-obese individuals.

***Classical Pooled Meta-analysis***

We next moved on to the pooled meta-analysis. For the Shannon diversity RR, all studies but Wu, Goodrich, and Turnbaugh had those with a lower Shannon diversity then the median as being higher risk for obesity [Figure 3A]. The pooled analysis showed an increased RR of `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['shannon', 'ci_lb'],2), nsmall=2)` - `r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`) for obesity in the lower then median Shannon diversity group that was significant (P-value = `r format(round(RR_composite['shannon', 'p_value']), nsmall=2)`). In contrast, when we investigated the B:F ratio RR  the pooled result from all studies showed that there was no significant difference between obese and non-obese individuals (P-value = `r format(round(RR_composite['bf_ratio', 'p_value']), nsmall=2)`) [Figure 3B].  The pooled RR was `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)` (CI `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)` - `r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`).  The differences in RR that was observed also do not seem to be related to the variable region chosen [Figure 3].

We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis. From the graph we can observe that there does not seem to be any significant  bias [Figure S5]. Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.

***Random Forest Classification and Predicition Accuracy***

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm. We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.  We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al. study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al. Study).  For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al. study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al. study). The total number of trees and node size was standardized to 500 and 20 respectively for each data set. Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05) regardless if it was using genus or OTU level classification [Figure 4].  Overall most studies performed poorly when used on other data sets for classification [Figure 5] with no significant difference between studies for their mean accuracy for prediction of non-obese and obese indivdiuals (P-value > 0.05).

***Power and Sample Size Estimates and Simulations***

For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 6A] compared to the B:F ratio [Figure 7A] differences.  When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 6A].  This predicted power was also similar for evenness (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`).   However, this was not the case for the B:F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 7A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.   Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 6B].  In contrast in order to detect a 5% difference with 80% power in the B:F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 7B].  A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S6-S11].

We also investigated if the combined total n for the linear mixed effect model was sufficiently powered.  Using a small (0.2), medium (0.5), and large (0.8) cohen value we estimated the power of the combined data sets as well as the average cohen value from across studies for Shannon Diversity (0.06) and BF ratio (0.01).  For the small effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.2, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.  For the medium effect for combined data power calculation a value of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.5, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.  For the large effect for combined data power calculation a value of
`r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=0.8, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`% was found.  The average actual cohens d value for Shannon Diversity for all `r length(rownames(beta_tests.summary))` studies of `r format(round(shannond, 2), nsmall=2)` gave a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=shannond, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.  Finally, the average actual cohens d value for B:F ratio for all `r length(rownames(beta_tests.summary))` studies of `r format(round(bf_ratiod, 2), nsmall=2)` yieled a power of `r format(round(pwr.t2n.test(n1=length(which(alpha.data.ALL$obese == FALSE)), n2=length(which(alpha.data.ALL$obese == TRUE)), d=bf_ratiod, sig.level = 0.05, alternative="two.sided")$power*100,2), nsmall=2)`%.

## Discussion:

@@

Fourth, although the ROC curves for each study were relatively good there was very little actual overlap between studies based on sequence similarity and directionality.  Further none of the selected variables from each indivdiual study did a respectable job of classification when applied to the other data sets.  This would suggest that within each indivdiual study classification of obese and non-obese is relatively strong but not generalizable across studies.  This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.  Additionally, an alternative explanation that may be more reasonable is that the random forest model is overfit for some of the studies examined which explains why the variables selected are less generalizable across studies.  In actuality it may be a combination of all of the factors mentioned as to why generalizability is problematic.

@@

**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B:F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for obese and non-obese groups.  When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 5), nsmall = 5)`). **B)** For the B:F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).

>>>>

# Lack of power and small effect size confounds the ability to differentiate non-obese and obese individuals using gut microbiome data

@@

```{r datasets}
alpha <- read.table(file="data/process/alpha_tests.summary", header=T)

alpha_composite <- read.table(file="data/process/alpha_composite.summary", header=T)
alpha_effect_size <- 100*abs(1-alpha_composite[,"obese"]/alpha_composite[,"non_obese"])
names(alpha_effect_size) <- alpha_composite$metric

relative <- read.table(file="data/process/relative_risk.summary", header=T)

relative_composite <- read.table(file="data/process/relative_risk.composite", header=T)

beta <- read.table(file="data/process/beta_tests.summary", header=T)

otu <- read.table(file="data/process/random_forest.otu.summary", header=T)

genus <- read.table(file="data/process/random_forest.genus.summary", header=T)

@@

rr_power <- read.table(file="data/process/rr_power.predicted", header=T)

@@

## Abstract
Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity. This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets. The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis. Alpha-diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances. Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%. A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity. When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`). Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.

@@

There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans. The first such study used genetically obese mice and observed the ratio of the relative abundances of Bacteroidetes to Firmicutes (B:F) was lower in obese mice than lean mice [@ley_obesity_2005]. Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009]. They also showed that the relative abundance of Bacteroidetes and Firmicutes increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006]. Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since direct manipulation of the communities yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016]. The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.

@@

```{r}
alpha_sig <- alpha[alpha$p.value < 0.05,]
sig_counts <- table(alpha_sig$metric)

stopifnot(sig_counts['bf_ratio']==0)
stopifnot(sig_counts['firmicutes']==0)
```

***Alpha diversity analysis***
We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample. Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the ten studies. The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study. We identified `r sum(alpha$p.value < 0.05)` P-values less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1). These results largely match those of the Walters and Finucane re-analysis studies. Interestingly, although only `r number_string[sig_counts["shannon"]]` of the 10 studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant. We used a random effects linear model to combine the studies using the study as the random effect and found statistical support for decreased richness, evenness, and diversity among obese individuals (all P<`r max(alpha_composite[alpha_composite$p <0.05, "p"])`). Although there was a significant relationship between these metrics and diversity and obesity status, the effect size was quite small. The obese individuals averaged `r alpha_effect_size["sobs"]`% lower richness, `r alpha_effect_size["shannoneven"]`% lower evenness, and `r alpha_effect_size["shannon"]`% lower diversity. There were no significant associations when we pooled the phylum-level metrics across studies. These results indicate that obese individuals do have a statistically significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant.

```{r}
rr_sig <- relative[relative$p.value < 0.05,]
sig_counts <- table(rr_sig$metric)

stopifnot(nrow(rr_sig) == 3)
stopifnot(sig_counts['bf_ratio']==0)
stopifnot(sig_counts['firmicutes']==0)
stopifnot(sig_counts['bacteroidetes']==0)

concatenate_rr <- function(data){

	paste0(format(data$rr, digits=2, nsmall=2L), " (", format(data$ci_lb, digits=2, nsmall=2L), "-", format(data$ci_ub, digits=2, nsmall=2L), ")")

}
```

***Relative risk***
Building upon the alpha diversity analysis we calculated the relative risk of being obese based on whether an individual's alpha diversity metrics were below or above the median metric for that study. The results using relative risk largely matched those of using the untransformed alpha diversity data. Across the ten studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2). Again, although the relative risk values were not significant for other studies, the values tended to be above one. When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`). The relative risks associated with alpha diversity were small. The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`. There were no significant difference in the phylum-level metrics. Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.

```{r}
beta_sig <- beta[beta$p_value < 0.05,]
stopifnot(beta_sig$dataset == c("escobar", "turnbaugh", "goodrich"))
```

***Beta diversity analysis***
Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study. We then used PERMANOVA to test for a significant differences between the structure of non-obese and obese individuals. The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05). Because it was not possible to ascertain the directionality of the difference in community structure nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.

```{r}
stopifnot(test_summary[which.min(test_summary$accuracy), "train"]=="turnbaugh")
stopifnot(test_summary[which.max(test_summary$accuracy), "train"]=="hmp")
```

***Development of a microbiome-based classifier of obesity***
The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The the observed AUC values for the models applied to the training datastes varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this to identify a threshold to calculate the accuracy of the models when applied to the other nine datasets (Figure 5). Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

```{r}
which_div_at_05 <- which(power[power$metric=="shannon" & power$effect_size==0.05,"power"] > 0.80)

stopifnot(power[which_div_at_05, "study"] == "goodrich")

which_div_at_10 <- which(power[power$metric=="shannon" & power$effect_size==0.10,"power"] > 0.80)

which_bf_at_15 <- which(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"] > 0.80)

stopifnot(length(which_bf_at_15) == 0)

stopifnot(sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.10,"power"] > 0.80) == 0)

range_effect_size <- range(abs(100*(1-alpha[alpha$metric == "shannon","mean_obese"]/alpha[alpha$metric=="shannon", "mean_non"])))

***Power and Sample Size Estimate Simulations***
The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals. To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S3-S8). Although there is no biological rationale for these effect sizes, they represent a range that is plausible. Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6). None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5). In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`. Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15. We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups. To detect a 1, 5, 10, and 15% difference in Shannon index, the median sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in B:F values, the median sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)`  individuals, respectively. To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric. In the case of obesity, the studies were underpowered to detect the `signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.

## Discussion

@@

Fourth, although the ROC curves for each study were relatively good there was very little actual overlap between studies based on sequence similarity and directionality.  Further none of the selected variables from each indivdiual study did a respectable job of classification when applied to the other data sets.  This would suggest that within each indivdiual study classification of non-obese and obese is relatively strong but not generalizable across studies.  This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.  Additionally, an alternative explanation that may be more reasonable is that the random forest model is overfit for some of the studies examined which explains why the variables selected are less generalizable across studies.  In actuality it may be a combination of all of the factors mentioned as to why generalizability is problematic.

@@

**Figure 2: Individual and Combined comparison of Obese and Non-Obese groups for Shannon Diversity and B:F Ratio.** **A)** For Shannon diversity only Baxter (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'baxter' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 2), nsmall = 2)` and Goodrich (P-value = `r format(round(alpha_tests.summary[which(alpha_tests.summary$dataset == 'goodrich' & alpha_tests.summary$metric == 'shannon'), 'p.value'], 5), nsmall = 5)`) studies indivdiually were signficantly different for non-obese and obese groups.  When the data sets were combined and study effect was corrected for,  there was a significant difference between Non-Obese and Obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 5), nsmall = 5)`). **B)** For the B:F ratio there was no individual study that was significant nor was their a difference when the data was combined and study effect corrected for between Non-Obese and Obese groups (P > 0.05).
